\Large{Learning to Pivot with Adversarial Networks}

### 2016-11-02 tex/paper.tex
Tex cleanup

%Many inference problems involve data generation processes that are not uniquely specified or uncertain in some way.
%In a scientific context, the presence of several plausible data generation processes is often
%associated to the presence of systematic uncertainty.
%We consider the situation of a
%family of data generation processes parametrized by nuisance parameters, which can
%be categorical or continuous. Robust inference is possible if it is based on a pivot -- a quantity whose
%distribution is invariant to the unknown value of these nuisance parameters.
%We introduce an adversarial training procedure to enforce the pivotal property and
%allow one to tune the tradeoff between power and robustness.

%Our goal is to have an inferential procedure that is robust to the uncertainties in the data generation process.
\end{abstract}

@@

% \kcnote{ need some text like abstract here before diving into formal statement.}

@@

\section{Problem statement}
\label{sec:problem}
%
%Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
%sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
%Let consider the multivariate random variables $X_z: \Omega \mapsto
%\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_z$ denotes a dependence
%of the functional $X$ on a nuisance parameter $Z$, whose values $z \in {\cal Z}$  define a
%parameterized family of its systematic uncertainties. That is, $X_z$ and $Y$
%induce together a joint probability distribution $p(X,Y|Z=z)$, where the
%conditional denotes $X_z$. For training, let further assume a finite set $\{
%x_i, y_i, z_i \}_{i=1}^N$ of realizations $X_{z_i}(\omega_i), Y(\omega_i)$, for
%$\omega_i \in \Omega$ and known values $z_i$ of the nuisance parameter. Our goal
%is to learn a score function $f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal S}$ of
%parameters $\theta_f$ (e.g., a neural network-based probabilistic classifier) and minimizing  a loss ${\cal L}_f(\theta_f)$ (e.g.,
%the cross-entropy). In addition, we require that $f(X_z ; \theta_f)$ should be
%robust to the value $z$ of the nuisance parameter  -- which remains unknown at
%test time. More specifically, we aim at building $f$ such that in the ideal case
%\begin{equation}\label{eqn:criterion-true}
%f(X_{z}(\omega) ; \theta_f) = f(X_{z^\prime}(\omega) ; \theta_f)
%\end{equation} for all
%samples $\omega \in \Omega$ and all $z, z^\prime$ pairs of values of the
%nuisance parameter.
%
%Since in general we do not have training tuples $(X_{z}(\omega),
%X_{z^\prime}(\omega))$ (for the same unknown $\omega$),

@@

%Note that a function
%$f$ for which Eqn.~\ref{eqn:criterion-true} is true necessarily satisfies
%Eqn.~\ref{eqn:criterion-measure}. In general, the converse is however not true,
%since the sets of samples $\{ \omega | f(X_{z}(\omega); \theta_f) = s \}$ and $\{
%\omega' | f(X_{z^\prime}(\omega'); \theta_f) = s \}$ do not need to be the same
%for the equality to hold. In order to simplify notations, and as only
%Eqn.~\ref{eqn:criterion-measure} is of direct interest in this work, we denote
%from here on the pivotal quantity criterion as

@@

%Experiments at high energy colliders like the LHC \citep{LHCMachine}
%are searching for evidence of new particles beyond the ones
%described by the so-called Standard Model (SM) of particle physics.
%A wide array of extended theories predict the existence of new massive
%particles that would decay to known heavy particles in the Standard Model,
%like the $W$, $Z$, Higgs boson, or top quark, which subsequently decay to
%multiple quarks.
%Finally, those quarks would produce collimated sprays of particles known as \textit{jets}.
%However, jets are produced ubiquitously at high energy colliders through
%more mundane processes in the SM, which leads to a challenging classification problem
%that is beset with a number of sources of systematic uncertainty.

Experiments at high energy colliders like the LHC \citep{LHCMachine}
are searching for evidence of new particles beyond the ones
described by the well established Standard Model (SM) of particle physics.
%Searches for these new particles in the LHC data involve a challenging
%classification problem that is beset with a number of sources of systematic uncertainty.
%
A wide array of extended theories predict the existence of new massive
particles that would decay to known particles in the SM such as the $W$ boson.
The $W$ boson is unstable and can decay to
two quarks, each of which produce collimated sprays of particles known as \textit{jets}
(collectively denoted $W\to jj$).
If the exotic particle is heavy, then the $W$ boson will be moving very fast,
and  relativistic effects will cause the two jets from its decay to
merge into a single `$W$-jet'.
These  $W$-jets have a rich internal substructure (see e.g.
\citep{Altheimer:2012mn,Altheimer:2013yza} and references therein).
However, jets are also produced ubiquitously at high energy colliders through
more mundane processes in the SM, which leads to a challenging classification problem
that is beset with a number of sources of systematic uncertainty.

%Distinguishing these ($y=1$) fat jets produced by heavy SM particles from
%the copious background of normal ($y=0$) jets is a fundamental challenge in searching
%for signs of new high mass particles.

%A wide array of extended theories predict the existence of new massive
%particles that would decay to known heavy particles in the Standard Model
%(e.g. the $W$ and $Z$ bosons), which subsequently decay to
%multiple quarks.
%Those quarks subsequently produce collimated sprays of particles known as \textit{jets} (denoted $j$).
%However, jets are produced ubiquitously at high energy colliders through
%more mundane processes in the SM, which leads to a challenging classification problem
%that is beset with a number of sources of systematic uncertainty.

%Consider the binary classification problem where $y=0$ corresponds to data produced from the
%ubiquitous SM process $pp \to jj$ and $y=1$ corresponds to data produced from an exotic
%process $pp \to W' \to W(\to jj) Z(\to jj)$ predicted by an extended theory.
%When the mass of the $W'$ is much larger than the mass of the $W$ and $Z$ bosons,
%then the $W$ and $Z$ particles are moving very fast and the relativistic Lorentz boost
%causes the two jets from their decay to merge into a single `fat jet',

%Consider the search for an exotic particle whose decay leads to a $W$ boson
%that subsequently decays to two jets via $W\to jj$.  If the $W$ boson is moving
%very fast, then the relativistic Lorentz boost
%causes the two jets from its decay to merge into a single `fat jet',
%which has a rich internal substructure (see e.g.
%\citep{Altheimer:2012mn,Altheimer:2013yza} and references therein).
%%We can cast this as a binary classification problem where $y=0$
%%corresponds to data produced from the ubiquitous SM process
%%that produce jets and $y=1$  corresponds to data produced from an exotic
%%process that gives rise to fat jets from $W\to jj$.

%binary classification problem where $y=0$ corresponds to data produced from the
%ubiquitous SM process $pp \to jj$ and $y=1$ corresponds to data produced from an exotic
%process $pp \to W' \to W(\to jj) Z(\to jj)$ predicted by an extended theory.
%When the mass of the $W'$ is much larger than the mass of the $W$ and $Z$ bosons,
%then the $W$ and $Z$ particles are moving very fast and the relativistic Lorentz boost
%causes the two jets from their decay to merge into a single `fat jet',
%which has a rich internal substructure (see e.g.
%\citep{Altheimer:2012mn,Altheimer:2013yza} and references therein).

%Distinguishing these ($y=1$) fat jets produced by heavy SM particles from
%the copious background of normal ($y=0$) jets is a fundamental challenge in searching
%for signs of new high mass particles.

@@

%Our goal is to build an accurate classifier, for which we also want to minimize the effects due to the
%uncertainties on the nuisance. More specifically, we choose to recast the
%classification problem as a hypothesis test between signal+background (jets
%originating from $W$ bosons or from quarks and gluons) and background
%only (jets originating from quarks and gluons only) and tune the decision
%threshold of our classifier by maximizing its approximate median significance
%(AMS), when uncertainties in the background are taken into account (see Eqn.~20
%of \citep{adam2014higgs}). Our motivation is that reducing the effects of
%uncertainties by requiring independence of $Z$ with the classifier output $f(X;\theta_f)$ should
%allow for a larger maximum significance.

@@

As  Fig.~\ref{fig:physics-ams} illustrates, without adversarial training (at
$\lambda=0|Z=0$ when building a classifier at the nominal value $Z=0$ only, or
at $\lambda=0$ when building a classifier on data sampled from $p(X,Y,Z)$), the
AMS peaks at $7$. By contrast, as the pivotal constraint
is made stronger (for $\lambda > 0$) the AMS peak moves higher, with a maximum
value around  $7.8$ for $\lambda=10$. In other words, trading classification
accuracy for robustness to pileup results in a net
benefit in terms of the power of the resulting hypothesis test. Setting $\lambda$ too high however
(e.g. $\lambda=500$) results in a decrease of the maximum AMS, by
focusing the capacity of $f$ too strongly on independence with $Z$, at the expense of
classification accuracy. As demonstrated in this example, optimizing $\lambda$
gives us a principled and effective approach to control the trade-off between
accuracy and robustness that ultimately maximizes the
power of the enveloping hypothesis test.
%significance by desensitizing
%the classifier output $f(X;\theta_f)$ in the most beneficial way.

@@

In machine learning, learning a pivotal quantity can be related to the problem
of domain adaptation
\citep{blitzer2006domain,pan2011domain,gopalan2011domain,gong2013connecting,baktashmotlagh2013unsupervised,ajakan2014domain,ganin2014unsupervised},
where the goal is often stated as trying to learn a domain-invariant
representation of the data. Likewise, our method also relates to the problem of
enforcing fairness in classification \citep{zemel2013learning,feldman2015certifying,EdwardsS15,zafar2015fairness}, which
is stated as learning a classifier that is independent of some chosen attribute
such as gender, color or age. For both families of methods, the problem can
equivalently be stated as learning a classifier which is a pivotal quantity with
respect to either the domain or the selected feature. In this context,
\citep{ganin2014unsupervised,EdwardsS15} are certainly among the closest to our
work, in which domain invariance and fairness are enforced through an
adversarial minimax setup composed of a classifier and an adversary
discriminator.
Following this line of work and without making assumptions on the prior $p(Z)$, our method can be regarded as a
generalization that also supports a continuously parametrized family of domains or as enforcing fairness over continuous attributes.
% In that sense, and to the best of our knowledge, our work is also original in the theoretical characterization
% of the minimax solution of Eqn.~\ref{eqn:min_thetaf}, as well as in how one can trade-off accuracy for robustness through Eqn.~\ref{eqn:vf-lambda}.

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty
 by estimating $p(f(x)|z)$ with a parametrized calibration
procedure. Clearly, this classifier is however not optimal for $z \neq z_0$.
To overcome this issue, the classifier $f$ is sometimes built instead on a mixture
of training data generated from several plausible values $z_0, z_1, \dots$ of the nuisance parameter.
While this certainly improves classification performance with respect to the marginal model $p(X,Y)$,
there is no reason to expect the resulting classifier to be pivotal, as shown
previously in Sec.~\ref{sec:toy}.
As an alternative, parametrized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, it is not obvious how such classifiers can be used on real data since
the correct value $z$ of the nuisance often remains unknown. This is
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
eliminated via optimization or marginalization.
%, but otherwise often limits the range of their applications.
In practice, parametrized classifiers  are also computationally expensive to build
and evaluate. In particular, calibrating their decision function, i.e.
approximating $p(f(x,z)|y,z)$ as a continuous function of $z$, remains an open
challenge. By contrast, constraining $f$ to be pivotal yields a classifier which
may not be optimal with respect to $Y|X,Z$, as discussed in
Sec.~\ref{sec:theory}, but that can otherwise be directly used in a wider range of
applications, since the dependence on the nuisance parameter $Z$ has already been eliminated.
Moreover, if $f$ is sufficiently close to being pivotal, then calibration only needs to be carried out once.

%Following this line of work and without making assumptions on the prior $p(Z)$, our method can be regarded as a
%generalization that also supports the continuous case, which can be viewed as
%handling infinitely many domains, provided they can be continuously
%parametrized, or as enforcing fairness over continuous attributes.

@@

\bigskip

>>>>

\end{abstract}

@@

@@

\section{Problem statement}
\label{sec:problem}

@@

@@

Experiments at high energy colliders like the LHC \citep{LHCMachine} are
searching for evidence of new particles beyond the ones described by the well
established Standard Model (SM) of particle physics. A wide array of extended
theories predict the existence of new massive particles that would decay to
known particles in the SM such as the $W$ boson. The $W$ boson is unstable and
can decay to two quarks, each of which produce collimated sprays of particles
known as \textit{jets} (collectively denoted $W\to jj$). If the exotic particle
is heavy, then the $W$ boson will be moving very fast, and  relativistic effects
will cause the two jets from its decay to merge into a single `$W$-jet'. These
$W$-jets have a rich internal substructure (see e.g.
\citep{Altheimer:2012mn,Altheimer:2013yza} and references therein). However,
jets are also produced ubiquitously at high energy colliders through more
mundane processes in the SM, which leads to a challenging classification problem
that is beset with a number of sources of systematic uncertainty.

@@

@@

As  Fig.~\ref{fig:physics-ams} illustrates, without adversarial training (at
$\lambda=0|Z=0$ when building a classifier at the nominal value $Z=0$ only, or
at $\lambda=0$ when building a classifier on data sampled from $p(X,Y,Z)$), the
AMS peaks at $7$. By contrast, as the pivotal constraint
is made stronger (for $\lambda > 0$) the AMS peak moves higher, with a maximum
value around  $7.8$ for $\lambda=10$. In other words, trading classification
accuracy for robustness to pileup results in a net
benefit in terms of the power of the resulting hypothesis test. Setting $\lambda$ too high however
(e.g. $\lambda=500$) results in a decrease of the maximum AMS, by
focusing the capacity of $f$ too strongly on independence with $Z$, at the expense of
classification accuracy. As demonstrated in this example, optimizing $\lambda$
gives us a principled and effective approach to control the trade-off between
accuracy and robustness that ultimately maximizes the
power of the enveloping hypothesis test.

@@

In machine learning, learning a pivotal quantity can be related to the problem
of domain adaptation
\citep{blitzer2006domain,pan2011domain,gopalan2011domain,gong2013connecting,baktashmotlagh2013unsupervised,ajakan2014domain,ganin2014unsupervised},
where the goal is often stated as trying to learn a domain-invariant
representation of the data. Likewise, our method also relates to the problem of
enforcing fairness in classification
\citep{zemel2013learning,feldman2015certifying,EdwardsS15,zafar2015fairness},
which is stated as learning a classifier that is independent of some chosen
attribute such as gender, color or age. For both families of methods, the
problem can equivalently be stated as learning a classifier which is a pivotal
quantity with respect to either the domain or the selected feature. In this
context, \citep{ganin2014unsupervised,EdwardsS15} are certainly among the
closest to our work, in which domain invariance and fairness are enforced
through an adversarial minimax setup composed of a classifier and an adversary
discriminator. Following this line of work and without making assumptions on the
prior $p(Z)$, our method can be regarded as a generalization that also supports
a continuously parametrized family of domains or as enforcing fairness over
continuous attributes.

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty by
estimating $p(f(x)|z)$ with a parametrized calibration procedure. Clearly, this
classifier is however not optimal for $z \neq z_0$. To overcome this issue, the
classifier $f$ is sometimes built instead on a mixture of training data
generated from several plausible values $z_0, z_1, \dots$ of the nuisance
parameter. While this certainly improves classification performance with respect
to the marginal model $p(X,Y)$, there is no reason to expect the resulting
classifier to be pivotal, as shown previously in Sec.~\ref{sec:toy}. As an
alternative, parametrized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, it is not obvious how such classifiers can be used on real
data since the correct value $z$ of the nuisance often remains unknown. This is
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
eliminated via optimization or marginalization. In practice, parametrized
classifiers  are also computationally expensive to build and evaluate. In
particular, calibrating their decision function, i.e. approximating
$p(f(x,z)|y,z)$ as a continuous function of $z$, remains an open challenge. By
contrast, constraining $f$ to be pivotal yields a classifier which may not be
optimal with respect to $Y|X,Z$, as discussed in Sec.~\ref{sec:theory}, but that
can otherwise be directly used in a wider range of applications, since the
dependence on the nuisance parameter $Z$ has already been eliminated. Moreover,
if $f$ is sufficiently close to being pivotal, then calibration only needs to be
carried out once.

@@
